---
title: "11_DataCleaning"
author: "Faith Ang, Ria Mathew, Xuan Wei Ding"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning

Data often needs to be cleaned. This ensures the best dataset to use for analysis.

```{r}
## everything we need plus the csv so it's not cluttered
source("00_requirements.R")

```

## Cleaning the Data

First, we will remove any rows with missing values. Then, we will convert categorical variables to factors.
This is done using functions defined in "01_funct_DataCleaning.R".
This sources functions from that file and applies them to the `hateCrime` dataset.

```{r, echo=FALSE}

source("01_funct_DataCleaning.R")
hateCrime_Clean <- remove_na(hateCrime)

hateCrime_Clean <- to_factors(hateCrime_Clean)
```

Now, we can view the cleaned dataset.
```{r}
head(hateCrime_Clean, 10)
```

The Question that we want to focus on is: 

"Which groups (racial, religious, gender, etc...) are most frequently targeted, and what events may have caused the increase or decrease of certain types of hate crime over the years?"

In order to answer this, it would be good to visualize the data and determine what can be gleaned from the data

Here are the list of variables available to us.
```{r}
colnames(hateCrime)
```

Of all of the variables, only few are relevant or useful to the study that will be conducted. These Variables are:

1. race of the offender : same
2. year : year of offense
3. state : state which offense took place
4. bias : group that hate crime was committed against
5. offense: what the offense type was (ie intimidation, actual assault) dictating the severity of the crime as well
6. victim count: how many victims per case

They are important because we want to draw comparisons between the variabes to draw conclusions that will lead us closer towards answering our overall questions. 

```{r}
hateCrime_renamed <- data_rename(hateCrime_Clean)
#create a new dataset with only those variables
hateCrime_Final <- hateCrime_renamed %>%
  select("Race of Offender", Offense, Bias, Year, State, "Victim Count")
# checking if the variables are named
hateCrime_Final
```

```{r}
# separating by state groups or regions

# west coast + hawaii and alaska??? 
hateCrime_Final$Region[hateCrime_Final$State == "California"] <- "West Coast"
hateCrime_Final$Region[hateCrime_Final$State == "Oregon"] <- "West Coast"
hateCrime_Final$Region[hateCrime_Final$State == "Washington"] <- "West Coast"

## non continental
hateCrime_Final$Region[hateCrime_Final$State == "Alaska"] <- "Non Continental"
hateCrime_Final$Region[hateCrime_Final$State == "Hawaii"] <- "Non Continental"
hateCrime_Final$Region[hateCrime_Final$State == "District of Columbia"] <- "Non Continental"

# federal

hateCrime_Final$Region[hateCrime_Final$State == "Federal"] <- "Federal"

# Northern east coast

hateCrime_Final$Region[hateCrime_Final$State == "Maine"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "Pennsylvania"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "Vermont"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "New Hampshire"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "Massachusetts"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "Connecticut"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "Rhode Island"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "New Jersey"] <- "Northeast"
hateCrime_Final$Region[hateCrime_Final$State == "New York"] <- "Northeast"

# southern US

hateCrime_Final$Region[hateCrime_Final$State == "Arizona"] <- "Southwest"
hateCrime_Final$Region[hateCrime_Final$State == "New Mexico"] <- "Southwest"
hateCrime_Final$Region[hateCrime_Final$State == "Oklahoma"] <- "Southwest"
hateCrime_Final$Region[hateCrime_Final$State == "Texas"] <- "Southwest"

hateCrime_Final$Region[hateCrime_Final$State == "Arkansas"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Missisippi"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Alabama"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Georgia"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "North Carolina"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "South Carolina"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Tennessee"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Kentucky"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Virginia"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "West Virginia"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Florida"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Louisiana"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Delaware"] <- "Southeast"
hateCrime_Final$Region[hateCrime_Final$State == "Maryland"] <- "Southeast"


# sw midwest

hateCrime_Final$Region[hateCrime_Final$State == "North Dakota"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "South Dakota"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Nebraska"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Kansas"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Minnesota"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Iowa"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Missouri"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Wisconsin"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Illinois"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Indiana"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Michigan"] <- "Midwest"
hateCrime_Final$Region[hateCrime_Final$State == "Ohio"] <- "Midwest"

# Rocky Mountians

hateCrime_Final$Region[hateCrime_Final$State == "Idaho"] <- "Rocky Mountains"
hateCrime_Final$Region[hateCrime_Final$State == "Utah"] <- "Rocky Mountains"
hateCrime_Final$Region[hateCrime_Final$State == "Wyoming"] <- "Rocky Mountains"
hateCrime_Final$Region[hateCrime_Final$State == "Colorado"] <- "Rocky Mountains"
hateCrime_Final$Region[hateCrime_Final$State == "Nevada"] <- "Rocky Mountains"
hateCrime_Final$Region[hateCrime_Final$State == "Montana"] <- "Rocky Mountains"

hateCrime_Final

```






