---
title: "11_DataCleaning"
author: "Faith Ang, Ria Mathew, Xuan Wei Ding"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning

Data often needs to be cleaned. This ensures the best dataset to use for analysis.

```{r}
## put the csv here
source("00_requirements.R")

#reads csv and names it for easy callback
hateCrime <- read_csv("hate_crime.csv")
```

## Cleaning the Data

First, we will remove any rows with missing values. Then, we will convert categorical variables to factors.
This is done using functions defined in "01_funct_DataCleaning.R".
This sources functions from that file and applies them to the `hateCrime` dataset.

```{r, echo=FALSE}

source("01_funct_DataCleaning.R")
hateCrime_Clean <- remove_na(hateCrime)

hateCrime_Final <- to_factors(hateCrime_Clean)
```

Now, we can view the cleaned dataset.
```{r}
head(hateCrime_Final, 10)
```

The Question that we want to focus on is: 

"Which groups (racial, religious, gender, etc...) are most frequently targeted, and what events may have caused the increase or decrease of certain types of hate crime over the years?"

In order to answer this, it would be good to visualize the data and determine what can be gleaned from the data

Of all of the variables, only few are relevant or useful to the study that will be conducted. These Variables are:

1. race of the offender : same
2. year : year of offense
3. state : state which offense took place
4. bias : group that hate crime was committed against
5. the offense (intimidation, actual assault, severity of the crime)

They are important because #insert why here#








