# for all ploting code and testing
As mentioned in the data cleaning file, we will be exploring the following prompts and basing the plots off of this.


1. count the number of offenses per bias to see which group is most targeted

Using a bar graph we can see that 


```{r}

# define race-related bias groups
race_groups <- c("Anti-Black or African American", 
                 "Anti-White", 
                 "Anti-Arab", 
                 "Anti-Asian", 
                 "Anti-Hispanic or Latino", 
                 "Anti-Multiple Races, Group", 
                 "Anti-Other Race/Ethnicity/Ancestry")

# filter rows where Bias is in your race_groups
race_data <- hateCrime_Final %>%
  filter(Bias %in% race_groups)

# count how many times each Bias appears
race_summary <- race_data %>%
  count(Bias, name = "total_incidents")

# plot
ggplot(race_summary, aes(x = reorder(Bias, -total_incidents), 
                         y = total_incidents, 
                         fill = Bias)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(title = "Total Number of Hate Crimes by Race",
       x = "Race Targeted",
       y = "Total Number of Reported Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}

# define sexuality based bias groups
sexuality_groups <- c("Anti-Gay (Male)", 
                      "Anti-Lesbian (Female)", 
                      "Anti-Heterosexual", 
                      "Anti-Lesbian, Gay, Bisexual, or Transgender (Mixed Group)")

# filter rows where Bias is in your race_groups
sex_data <- hateCrime_Final %>%
  filter(Bias %in% sexuality_groups)

# count how many times each Bias appears
sex_summary <- sex_data %>%
  count(Bias, name = "total_incidents")

# plot
ggplot(sex_summary, aes(x = reorder(Bias, -total_incidents), 
                         y = total_incidents, 
                         fill = Bias)) +
  geom_bar(stat = "Identity") +
  labs(title = "Total Number of Hate Crimes by Sexuality",
       x = "Sexuality Targeted",
       y = "Total Number of Reported Incidents") 

```



2. count the number of offenses per race to see which group is more likely to be the offender

```{r}

offender_race 

```

3. how has the number of offenses changed over the years 

```{r}

```

4. variation by state of types of offenses

```{r}

```

5. how many of each type of offense there is and compare

```{r}

```

6. how biases and offense correlate (eg. do asians experience intimidation based hate crimes or assult based hate crimes)
